---
title: "HW 1 Solutions"
author: "Hakan Gogtas"
date: "9/22/2019"
output: pdf_document
---


```{r setup, include=FALSE}

```

## Question 1 

_1-) The regression model we would like to study is:_

$$ Y_i = \beta_0  + \varepsilon_i$$

a-) the least square estimation for $$\beta_0$$


$$Q=\sum(Y_i - Yhat_i)^2 = \sum (Y_i - \beta_0)^2 $$
$$@Q/\beta_0=-2\sum (Y_i - \beta_0)= 0$$
$$\beta_0 = \frac{sum(Y_i)}{n}=Ybar$$
$$ \dots= E(exp(\beta_0 + \beta_1X_{i1} + \beta_2X_{i2})) + E\varepsilon_i - E\delta_i = exp (\beta_0 + \beta_1X_{i1} + \beta_2X_{i2})  - \alpha X_{i1} X_{i2} $$


## Question 2

_The dataset teengamb (see below for the instructions in r) concerns a study of teenage gambling in Britain. Make a numerical and graphical summary of the data, commenting on any features that you find interesting. Limit the output you present to a quantity that a busy reader would find sufficient to get a basic understanding of the data.   

library(faraway)                             # download the library
data.help("teengamb")                 # see the description of the data


```{r q2a}
library(faraway) 
data("teengamb")
plot(teengamb)
attach(teengamb)
par(mfrow=c(1,4))
boxplot(status~sex,data=teengamb, main="Gender Impact", xlab="Sex", ylab="Status")
boxplot(income~sex,data=teengamb, main="Gender Impact", xlab="Sex", ylab="Income")
boxplot(verbal~sex,data=teengamb, main="Gender Impact", xlab="Sex", ylab="Verbal")
boxplot(gamble~sex,data=teengamb, main="Gender Impact", xlab="Sex", ylab="Gamble")

cor(teengamb)

```

There is a difference in sex for status and gamble. There is one outlier for Gamble, and one for Income. Income and Gamble are significantly correlated. 

## Question 3
_3.)Refer to the CDI data set. The number of active physicians in a CDI (Y) is expected to be related to total population, number of hospital beds, and total personal income. 

a. Regress the number of active physicians in turn on each of the three predictor variables. State the estimated regression functions.
b. Plot the three estimated regression functions and data on separate graphs. Does a linear regression relation appear to provide a good fit for each of the three predictor variables?
c. Calculate MSE for each of the three predictor variables. Which predictor variable leads to the smallest variability around the fitted regression line?

```{r q3}
f1<-lm(`Number of active physicians` ~ `Total population`,data=CDI)
f2<-lm(`Number of active physicians` ~ `Number of hospital beds`,data=CDI)
f3<-lm(`Number of active physicians` ~ `Total personal income`,data=CDI)
par(mfrow=c(1,3))
plot(CDI$`Total population`,CDI$`Number of active physicians`)
abline(f1)
plot(CDI$`Number of hospital beds`,CDI$`Number of active physicians`)
abline(f2)
plot(CDI$`Total personal income`,CDI$`Number of active physicians`)
abline(f3)
anova(f1)
anova(f2)
anova(f3)

```

Y= -1.106e+02 + 2.795e-03*Total Population.  MSE=372204
Y= -95.9322 + 0.7431*Number of hospital beds.MSE=310192
Y= -48.3948 +  0.1317*Total personal income. MSE=324539

the linear relationship provide a good fit. Number of hospital beds has te highest preditibility power
## Question 4

```{r q4}
set.seed(12345)
ind <- sample(1:nrow(CDI), size =nrow(CDI)*0.70)
dev <- CDI[ind,]
holdout <- CDI[-ind,]
f11<-lm(`Number of active physicians` ~ `Total population`,data=dev)
f21<-lm(`Number of active physicians` ~ `Number of hospital beds`,data=dev)
f31<-lm(`Number of active physicians` ~ `Total personal income`,data=dev)
par(mfrow=c(1,3))
plot(dev$`Total population`,dev$`Number of active physicians`)
abline(f11)
plot(dev$`Number of hospital beds`,dev$`Number of active physicians`)
abline(f21)
plot(dev$`Total personal income`,dev$`Number of active physicians`)
abline(f31)
anova(f11)
anova(f21)
anova(f31)
MSE1<-sum((holdout$`Number of active physicians`-predict(f11,holdout))^2)/( dim(holdout)[1]-2)
MSE2<-sum((holdout$`Number of active physicians`-predict(f21,holdout))^2)/( dim(holdout)[1]-2)
MSE3<-sum((holdout$`Number of active physicians`-predict(f31,holdout))^2)/( dim(holdout)[1]-2)
cbind(MSE1,MSE2,MSE3)
```

total personel income has the highest predictibility power based on MSE calculated on the holdout sample.

## Question 5


```{r q5}
f<-lm(gamble~income,data=teengamb)
summary(f)
summary(f$residuals)
res<-f$residuals
res<-cbind(f$model,res)
res[which.min(res$res),]
res[which.max(res$res),]
```

_Gamble=-6.325 + 5.520*income 
_summary(f$residuals)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-46.020 -11.874  -3.757   0.000  11.934 107.120 

24th observation has the largest redisudual  
> res[which.min(res$res),]
   gamble income       res
17    0.1    9.5 -46.02005
> res[which.max(res$res),]
   gamble income      res
24    156     10 107.1197

