---
title: 'CS-E-106: Data Modeling - Midterm Exam'
subtitle: '**Question 4**'
date: '**Due Date:** 10/21/2019'
author:
  - '**Instructor: Hakan Gogtas**'
  - '**Submitted by:** Saurabh Kulkarni'
output:
  pdf_document: 
    latex_engine: xelatex
---


**Solution 4:**

```{r}
q4_data = read.csv("question4.csv")
lm_q4 = lm(Y~X, data=q4_data)
summary(lm_q4)

```

The regression function: $Y = 156.3466 + 1.1900*X$

```{r}
build_residual_qq(lm=lm_q4, df=q4_data, rse=8.173)
```


*Interpretation:*

*Fitted vs. Residual Plot:* The residual plot appears to be equally spread and has no distinct patterns and no visible extreme outliers. We can say that there is mostly a contant variance in the error term.

*Normal Probability Plot:* The plot is mostly linear, which means that the error is in agreement with the normality.


**(B)**

*Breusch-Pagan Test*

Null Hypothesis: $H_{0}$: Error variance is constant
Alternate Hypothesis: $H_{1}$: Error variance is not constant

```{r}
ei = lm_q4$residuals
ei2 = ei^2
f = lm(ei2~q4_data$X)
summary(f)

#to find SSE(R) and SSR(R)
anova(f)

#to find SSE(F) and SSR(F)
anova(lm_q4)
```


```{r}
SSR_R = 31833 
SSE_R = 400089 

SSR_F = 11627.5
SSE_F= 3874.4

n = nrow(q4_data)

#chi-squared: [SSR(R)/2] / [SSE(F)/n]^2  
chiTest = (SSR_R/2) / ((SSE_F/n))^2
print(chiTest)


#p 
chi = qchisq(1-0.05,1)
print(chi)

```

Decision Rule:

- If $chiTest \leq \chi^{2}(1-\alpha,1)$, conclude $H_{0}$: constant error variance

- If $chiTest > \chi^{2}(1-\alpha,1)$, conclude $H_{1}$: non-constant error variance 


Result:
Since $3.817167 \leq 3.841459$ i.e. $chiTest \leq \chi^{2}(1-\alpha,1)$, we conclude $H_{0}$. The error variance is constant.

```{r}
```


```{r}
```


```{r}
```


```{r}
```

