---
title: '**CS-E-106: Data Modeling**'
subtitle: '**Assignment 6**'
date: '**Due Date:** 11/11/2019'
author:
  - '**Instructor: Hakan Gogtas**'
  - '**Submitted by:** Saurabh Kulkarni'
output: pdf_document
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), 
                      fig.width=10, fig.height=5)

loadLib = function(libName)
{
    if(require(libName, character.only=TRUE))
    {
        cat(libName, "loaded properly\n")
    } else {
        cat("Installing", libName, "\n")
        install.packages(libName)
        if(require(libName, character.only=TRUE))
        {
            cat(libName, "loaded properly\n")
        } else {
            stop(c(libName, "not properly installed\n"))
        }
    }
}

libs = c("ggplot2", "knitr", "MASS", "ggcorrplot", "GGally", "alr3", "lattice")

for (lib in libs)
{
    loadLib(lib)
}

```



**Question 1:**

1-	An analyst wanted to fit the regression model $Y_i = \beta_0 +\beta_1*X_{i1}+\beta_2*X_{i2}+\beta_1*X_{i3}+\epsilon_i$,
i = 1, ... , n, by the method of least squares when it is known that $\beta_2=4$. How can the analyst obtain the desired fit by using a multiple regression computer program?

**Solution 1:**


*Step 1:* Create a 



**Solution 2:**

**(a)** Obtain the analysis of variance table that decomposes the regression sum of squares into extra sums of squares associated with X4; with X1 given X4; with X2 , given X1 and X4; and with X3 , given X1, X2 and X4. (10pts)

```{r}
properties_data = read.csv("Commercial Properties.csv")
lm_prop = lm(Y~X4+X1+X2+X3, data=properties_data)
summary(lm_prop)
```


```{r}
anova_F = anova(lm_prop)
anova_F

anovaTable = data.frame(anova_F)
totals = c(round(sum(anovaTable[,1])), round(sum(anovaTable[,2])), "", "", "")
anovaTable = rbind(anovaTable, totals)

#add names to the table 
row.names(anovaTable) = c("SSR(X4)", "SSR(X1|X4)", "SSR(X2|(X4X1))","SSR(X3|(X4X1X2))", "SSE", "Total")
colnames(anovaTable) = c("DF", "Sum Sq.", "Mean Sq.", "F-Value", "Pr(>F)")

kable(anovaTable)

```


**(b)**	

From the above ANOVA table, we can see that the P-value for SSR(X3|X4X1X2) is very high, which means that the extra regression sums of squares due to X3 is very low. Thus, X3 can be dropped. F-test below.


```{r}
ssr = as.numeric(anovaTable["SSR(X3|(X4X1X2))","Sum Sq."])
sse = as.numeric(anovaTable["SSE","Sum Sq."])
df_diff = 1
df_E = as.numeric(anovaTable["SSE","DF"])

FStar = (ssr/df_diff) / (sse/df_E)
print(FStar)

print(paste("P-value:", 1-pf(FStar, df_diff, df_E)))

#alpha is given
alpha = 0.01

# df from Summary above in a
FTest = qf(1-alpha, df_diff, df_E)
print(FTest)
```


*Hypotheses:*

$H_0: \beta_3 = 0$ 

$H_a: \beta_3 \neq 0$ 


*Decision Rules:*

If $F^* \leq$ `r FTest`, conclude $H_0$

If $F^* >$ `r FTest`, conclude $H_a$

*Conclusion:*

Since our test statistic, $F^* =$ `r FStar`, and `r FStar` $\leq$ `r FTest`, we conclude $H_0$. Thus, X3 can be dropped from the model.


**(c)**	Test whether both X2 and X3 can be dropped from the regression model given that X1 and X4 are retained; use α=.01. State the alternatives, decision rule, and conclusion. What is the P-value of the test? (5pts)



```{r}
ssr = sum(as.numeric(anovaTable[c("SSR(X3|(X4X1X2))","SSR(X2|(X4X1))"),"Sum Sq."]))
sse = as.numeric(anovaTable["SSE","Sum Sq."])

df_diff = 2
df_E = as.numeric(anovaTable["SSE","DF"])

FStar = (ssr/df_diff) / (sse/df_E)
print(FStar)

print(paste("P-value:", 1-pf(FStar, df_diff, df_E)))

#alpha is given
alpha = 0.01

# df from Summary above in a
FTest = qf(1-alpha, df_diff, df_E)
print(FTest)

```

*Hypotheses:*

$H_0: \beta_2 = \beta_3 = 0$ 

$H_a:$ Not both $\beta$s equal to zero


*Decision Rules:*

If $F^* \leq$ `r FTest`, conclude $H_0$

If $F^* >$ `r FTest`, conclude $H_a$

*Conclusion:*

Since our test statistic, $F^* =$ `r FStar`, and `r FStar` $>$ `r FTest`, we conclude $H_1$. Not both $\beta$s equal to zero.


**(d)**


```{r}
Y_new = properties_data$Y+0.1*properties_data$X1-0.4*properties_data$X2
lm_prop_R = lm(Y_new~properties_data$X3+properties_data$X4)
summary(lm_prop_R)
```


```{r}
anova_R = anova(lm_prop_R)
anova_R
```


```{r}
SSE_R = anova_R["Residuals", "Sum Sq"]
SSE_F = anova_F["Residuals", "Sum Sq"]
df_R = anova_R["Residuals", "Df"]
df_F = anova_F["Residuals", "Df"]

FStar = ((SSE_R-SSE_F)/(df_R-df_F))/(SSE_F/df_F)
print(FStar)

#alpha is given
alpha = 0.01

# df from Summary above in a
FTest = qf(1-alpha, (df_R-df_F), df_F)
print(FTest)
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```






