---
title: '**CS-E-106: Data Modeling**'
subtitle: '**Assignment 6**'
date: '**Due Date:** 11/11/2019'
author:
  - '**Instructor: Hakan Gogtas**'
  - '**Submitted by:** Saurabh Kulkarni'
output: 
  pdf_document: 
    latex_engine: xelatex
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), 
                      fig.width=10, fig.height=5)

loadLib = function(libName)
{
    if(require(libName, character.only=TRUE))
    {
        cat(libName, "loaded properly\n")
    } else {
        cat("Installing", libName, "\n")
        install.packages(libName)
        if(require(libName, character.only=TRUE))
        {
            cat(libName, "loaded properly\n")
        } else {
            stop(c(libName, "not properly installed\n"))
        }
    }
}

libs = c("ggplot2", "knitr", "MASS", "ggcorrplot", "GGally", "alr3", "lattice", "dplyr")

for (lib in libs)
{
    loadLib(lib)
}

```



**Question 1:**

1-	An analyst wanted to fit the regression model $Y_i = \beta_0 +\beta_1*X_{i1}+\beta_2*X_{i2}+\beta_1*X_{i3}+\epsilon_i$,
i = 1, ... , n, by the method of least squares when it is known that $\beta_2=4$. How can the analyst obtain the desired fit by using a multiple regression computer program?

**Solution 1:**


*Step 1:* Create a new variable $Y^* = Y_i-\beta_2*X_{i2} = Y_i-4*X_{i2}$
*Step 2:* Using $Y^*$ as the new response variable, run the regression model: 
$Y^* = \beta_0 +\beta_1*X_{i1}+\beta_1*X_{i3}+\epsilon_i$ using least squares (Using R functions: lm(YStar~X1+X3, data)).
*Step 3:* Use the obtained coefficients as $\beta_1$ and $\beta_3$, assuming $\beta_2=4$



**Solution 2:**

**(a)** Obtain the analysis of variance table that decomposes the regression sum of squares into extra sums of squares associated with X4; with X1 given X4; with X2 , given X1 and X4; and with X3 , given X1, X2 and X4. (10pts)

```{r}
properties_data = read.csv("Commercial Properties.csv")
lm_prop = lm(Y~X4+X1+X2+X3, data=properties_data)
summary(lm_prop)
```


```{r}
anova_F = anova(lm_prop)
anova_F

anovaTable = data.frame(anova_F)
totals = c(round(sum(anovaTable[,1])), round(sum(anovaTable[,2])), "", "", "")
anovaTable = rbind(anovaTable, totals)

#add names to the table 
row.names(anovaTable) = c("SSR(X4)", "SSR(X1|X4)", "SSR(X2|(X4X1))","SSR(X3|(X4X1X2))", "SSE", "Total")
colnames(anovaTable) = c("DF", "Sum Sq.", "Mean Sq.", "F-Value", "Pr(>F)")

kable(anovaTable)

```


**(b)**	

From the above ANOVA table, we can see that the P-value for SSR(X3|X4X1X2) is very high, which means that the extra regression sums of squares due to X3 is very low. Thus, X3 can be dropped. F-test below.


```{r}
ssr = as.numeric(anovaTable["SSR(X3|(X4X1X2))","Sum Sq."])
sse = as.numeric(anovaTable["SSE","Sum Sq."])
df_diff = 1
df_E = as.numeric(anovaTable["SSE","DF"])

FStar = (ssr/df_diff) / (sse/df_E)
print(FStar)

print(paste("P-value:", 1-pf(FStar, df_diff, df_E)))

#alpha is given
alpha = 0.01

# df from Summary above in a
FTest = qf(1-alpha, df_diff, df_E)
print(FTest)
```


*Hypotheses:*

$H_0: \beta_3 = 0$ 

$H_a: \beta_3 \neq 0$ 


*Decision Rules:*

If $F^* \leq$ `r FTest`, conclude $H_0$

If $F^* >$ `r FTest`, conclude $H_a$

*Conclusion:*

Since our test statistic, $F^* =$ `r FStar`, and `r FStar` $\leq$ `r FTest`, we conclude $H_0$. Thus, X3 can be dropped from the model.


**(c)**	Test whether both X2 and X3 can be dropped from the regression model given that X1 and X4 are retained; use α=.01. State the alternatives, decision rule, and conclusion. What is the P-value of the test? (5pts)



```{r}
ssr = sum(as.numeric(anovaTable[c("SSR(X3|(X4X1X2))","SSR(X2|(X4X1))"),"Sum Sq."]))
sse = as.numeric(anovaTable["SSE","Sum Sq."])

df_diff = 2
df_E = as.numeric(anovaTable["SSE","DF"])

FStar = (ssr/df_diff) / (sse/df_E)
print(FStar)

print(paste("P-value:", 1-pf(FStar, df_diff, df_E)))

#alpha is given
alpha = 0.01

# df from Summary above in a
FTest = qf(1-alpha, df_diff, df_E)
print(FTest)

```

*Hypotheses:*

$H_0: \beta_2 = \beta_3 = 0$ 

$H_a:$ Not both $\beta$s equal to zero


*Decision Rules:*

If $F^* \leq$ `r FTest`, conclude $H_0$

If $F^* >$ `r FTest`, conclude $H_a$

*Conclusion:*

Since our test statistic, $F^* =$ `r FStar`, and `r FStar` $>$ `r FTest`, we conclude $H_1$. Not both $\beta$s equal to zero.


**(d)**


```{r}
Y_new = properties_data$Y+0.1*properties_data$X1-0.4*properties_data$X2
lm_prop_R = lm(Y_new~properties_data$X3+properties_data$X4)
summary(lm_prop_R)
```


```{r}
anova_R = anova(lm_prop_R)
anova_R
```


```{r}
SSE_R = anova_R["Residuals", "Sum Sq"]
SSE_F = anova_F["Residuals", "Sum Sq"]
df_R = anova_R["Residuals", "Df"]
df_F = anova_F["Residuals", "Df"]

FStar = ((SSE_R-SSE_F)/(df_R-df_F))/(SSE_F/df_F)
print(FStar)

#alpha is given
alpha = 0.01

# df from Summary above in a
FTest = qf(1-alpha, (df_R-df_F), df_F)
print(FTest)
```

**Solution 3:**

**(a)**

```{r}
standardize_corr = function(df){
  cols = colnames(df)
  df_new = df
  n = nrow(df)
  for(c in cols){
    mu = mean(df[, c])
    s = sqrt(var(df[, c]))
    df_new[, c] = (df[, c]-mu)/(s*sqrt(n-1))
  }
  df_new
}
```

```{r}
brand_data = read.csv("Brand Preference.csv")
brand_data_new = standardize_corr(brand_data)
summary(brand_data_new)
```

```{r}
lm_brand_new = lm(Y~., data=brand_data_new)
summary(lm_brand_new)
```

$\beta_0$ is almost 0. Thus,

Regression Function: $Y = 0.892*X_1 + 0.3946*X_2$

```{r}
lm_brand = lm(Y~., data=brand_data)
```


```{r}
library(QuantPsyc)
lm.beta(lm_brand)
```

**(b)**

*Interpretation:*

- We can see that the $\beta_0$ is almost equal to 0, which is expected since Y is now centered at 0 (based on definition and summary of the new data in part (a)).
- 

**(c)**


```{r}
lm_brand$coefficients
```


**(d)**


```{r}
df = brand_data

R2_Y1 = anova(lm(Y~X1, data=df))[1,2]/sum(anova(lm(Y~X1, data=df))[1:2,2])

R2_Y2 = anova(lm(Y~X2, data=df))[1,2]/sum(anova(lm(Y~X2, data=df))[1:2,2])

R2_12 = sum(anova(lm(Y~X1+X2, data=df))[1:2,2])/sum(anova(lm(Y~X1+X2, data=df))[1:3,2])

R2_Y1_2 = anova(lm(Y~X2+X1, data=df))[2,2]/sum(anova(lm(Y~X2+X1, data=df))[2:3,2])

R2_Y2_1 = anova(lm(Y~X1+X2, data=df))[2,2]/sum(anova(lm(Y~X1+X2, data=df))[2:3,2])

R2 = R2_12

```

$R_{Y1}^2 =$ `r R2_Y1`

$R_{Y2}^2 =$ `r R2_Y2`

$R_{12}^2 =$ `r R2_12`

$R_{Y1|2}^2 =$ `r R2_Y1_2`

$R_{Y2|1}^2 =$ `r R2_Y2_1`

$R^2 =$ `r R2`

**Solution 4:**

**(a)**

```{r}
inc_cols = c("Number.of.active.physicians", "Total.population", "Total.personal.income",
             "Land.area", "Percent.of.population.65.or.older", "Number.of.hospital.beds", 
             "Total.serious.crimes")
cdi_data = read.csv("CDI.csv")[, inc_cols]
colnames(cdi_data)
```



```{r}
cdi_data = 
cdi_data %>%
  rename(
    Y = Number.of.active.physicians,
    X1 = Total.population,
    X2 = Total.personal.income,
    X3 = Land.area,
    X4 = Percent.of.population.65.or.older, 
    X5 = Number.of.hospital.beds,
    X6 = Total.serious.crimes
  )

colnames(cdi_data)
```


```{r}
lm_cdi = lm(Y~X1+X2, data=cdi_data)
summary(lm_cdi)
```


```{r}
df = cdi_data

print(anova(lm(Y~X1+X2+X3,df)))
(r2_X3 = anova(lm(Y~X1+X2+X3,df))[3,2]/sum(anova(lm(Y~X1+X2+X3,df))[3:4,2]))
print(anova(lm(Y~X1+X2+X4,df)))
(r2_X4 = anova(lm(Y~X1+X2+X4,df))[3,2]/sum(anova(lm(Y~X1+X2+X4,df))[3:4,2]))
print(anova(lm(Y~X1+X2+X5,df)))
(r2_X5 = anova(lm(Y~X1+X2+X5,df))[3,2]/sum(anova(lm(Y~X1+X2+X5,df))[3:4,2]))
print(anova(lm(Y~X1+X2+X6,df)))
(r2_X6 = anova(lm(Y~X1+X2+X6,df))[3,2]/sum(anova(lm(Y~X1+X2+X6,df))[3:4,2]))

```

$R_{3|12}^2 =$ `r r2_X3`

$R_{4|12}^2 =$ `r r2_X4`

$R_{5|12}^2 =$ `r r2_X5`

$R_{6|12}^2 =$ `r r2_X6`


**(b)**

$X_5$ is the best predictor we can add to the model as it has the maximum coefficient of partial determination. Yes, the extra sum of squares associated with this variable is larger compared to other variables also, which makes sense since SST will remain constant.


**(c)**

```{r}
anova_x5 = anova(lm(Y~X1+X2+X5, data=cdi_data))
anova_x5
```


```{r}
ssr = as.numeric(anova_x5["X5","Sum Sq"])
sse = as.numeric(anova_x5["Residuals","Sum Sq"])
df_diff = 1
df_E = as.numeric(anova_x5["Residuals","Df"])

FStar = (ssr/df_diff) / (sse/df_E)
print(FStar)

print(paste("P-value:", 1-pf(FStar, df_diff, df_E)))

#alpha is given
alpha = 0.01

# df from Summary above in a
FTest = qf(1-alpha, df_diff, df_E)
print(FTest)

```

*Hypotheses:*

$H_0: \beta_5 = 0$ 

$H_a: \beta_5 \neq 0$ 


*Decision Rules:*

If $F^* \leq$ `r FTest`, conclude $H_0$

If $F^* >$ `r FTest`, conclude $H_a$

*Conclusion:*

Since our test statistic, $F^* =$ `r FStar`, and `r FStar` $\leq$ `r FTest`, we conclude $H_0$. Thus, X3 can be dropped from the model.



```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```




