---
title: "Practice Midterm Solutions"
output:
  pdf_document: default
  html_document: default
---
## Question 1 

_1-) The regression model we would like to study is:_

$$ Y_i =  \beta_1X_{i} + \varepsilon_i$$

_a-)write the likelihood function_


$$L =  \prod_{i=1}^n \frac{1}{\sqrt{2\pi\sigma^2}}exp\left(-\frac{(Y_i-\beta_1X_{i})^2}{2\sigma^2}\right) $$

_a-)find the MLS estimations for b_1 and $\sigma^2$_
_log likelihood function_ 

$$Log L = -\frac{n}{2}log2\pi-\frac{n}{2}log\sigma^2 -\frac{1}{2\sigma^2} \sum_{i=1}^n (Y_i-\beta_1X_{i})^2 $$

$$ \frac{\partial L}{\partial \beta_1}=\sum_{i=1}^n (Y_iX_i-\beta_1X^2_{i})=0 $$
$$\beta_1= \frac{\sum_{i=1}^n Y_iX_i}{\sum_{i=1}^nX^2_{i}} $$
$$ \frac{\partial L}{\partial \sigma}=-\frac{n}{\sigma}+\frac{1}{\sigma^3}\sum_{i=1}^n (Y_i-\beta_1X_{i})^2=0 $$
$$ \sigma^2=\frac{\sum_{i=1}^n (Y_i-\beta_1X_{i})^2}{n}$$

## Problem 2
### n	 Xbar	Ybar	SXX	 SYY	SXY	SSE
###120	28	3	    2340	49	92	46

_a-)	calculate b_0 and b_1 and 95% confidence intervals for b_0 and b_1_

### _Solution:_

```{r}
library(knitr)
n=120
Xbar=28
Ybar=3
SXX=2340
SYY=49
SXY=92
SSE=46

b1= SXY/SXX  
b0 =  Ybar-b1*Xbar
MSE = SSE/(n-2)
Var.b1 = MSE/SXX
Var.bo= MSE*{1/n + Xbar^2 / SXX}
sb1 = sqrt(Var.b1)
sb0 = sqrt(Var.bo)
tc=qt(1-0.05/2,n-2)
cbind(b1-tc*sb1,b1+tc*sb1)
cbind(b0-tc*sb0,b0+tc*sb0)


```
_b)	for Xh=3, predict Yh and calculate 95% prediction interval_
```{r}

Yhat=b0+b1*3
Yhat
s.yhat=sqrt(MSE*((1/n)+((3-Xbar)^2/SXX)))
cbind(Yhat-tc*s.yhat,Yhat+tc*s.yhat)
```
## Problem 3
### 3.	Based on the table above, write the ANOVA table and perform the F test, and perform a General Linear F test?



```{r}
SST = SYY
SSR= SST - SSE
cbind(SSR,SSE,SST)
Df.r=1
Df.e=n-2
cbind(1,n-2,n-1)
MSR=SSR
MSE=SSE/(n-2)
Ftest=MSR/MSE
cbind(MSR,MSE)
Ftest
1-pf(Ftest,1,n-2)


```
# Source         DF    SS   MS    F    
# Regression      1     3   3
# Error         118    46   0.39  7.7
# Total         119    49
# Ho:B_1=0
# Ha:B_1
# F = (49-46)/(46/118) = 3/0.39=7.7, pavalue=0.006, reject null, model is significant.

## Problem 4

### a)	Fit a regression model to predict house price. Obtain, the residuals and plot them against the fitted values. Also prepare a normal probability plot. What do your plots show?
### b)	Conduct the Brown-Forsythe test to determine whether or not the error variance varies with the level of X.
### c)	Calculate the simultaneous interval for  b_o,and b_1
### d)	Calculate the simultaneous confidence intervals for the predicted house prices for 1200,1400,1500 square feet

### _Solution:_
### a)

```{r}
House.Price<-c(245,312,279,308,199,219,405,324,319,255)
Square.Feet<-c(1400,1600,1700,1875,1100,1550,2350,2450,1425,1700)
f4<-lm(House.Price~Square.Feet)
summary(f4)
ei<-f4$residuals
yhat=f4$fitted.values
stdei<- rstandard(f4)
par(mfrow=c(1,2))
plot(ei,yhat,xlab="Residuals",ylab="Fitted Values")
qqnorm(stdei,ylab="Standardized Residuals",xlab="Normal Scores", main="QQ Plot") 
qqline(stdei,col = "steelblue", lwd = 2)

```

### The QQ plot roughly looks normal, error vs fitted values do not indicate unequal variance.


### b)	Conduct the Brown-Forsythe test to determine whether or not the error variance varies with the level of X.
```{r}
ei<-f4$residuals
M=median(Square.Feet)
DM<-data.frame(cbind(House.Price,Square.Feet,ei))
DM1<-DM[DM[,2]< M,]
DM2<-DM[DM[,2]>=M,]

M1<-median(DM1[,3])
M2<-median(DM2[,3])
N1<-length(DM1[,3])
N2<-length(DM2[,3])

d1<-abs(DM1[,3]-M1)
d2<-abs(DM2[,3]-M2)
s2<-sqrt((var(d1)*(N1-1)+var(d2)*(N2-1))/(N1+N2-2))
Den<- s2*sqrt(1/N1+1/N2)
Num<- mean(d1)-mean(d2)
T= Num/Den
T

```
### t stat is less than 1, indicates that null shouls be accepted. There is No unequal variance.
### c)

```{r}
confint(f4,level=1-0.05/2)

```
###d)	Calculate the simultaneous confidence intervals for the predicted house prices for 1200,1400,1500 square feet


```{r}
Xh<-c(1200,1400,1500)
predict.lm(f4,data.frame(Square.Feet  = c(Xh)),interval = "confidence", level = 1-0.05/3)
```

## Problem 5

### a)	Fit a linear regression function. Obtain, the residuals and plot them against the fitted values. Also prepare a normal probability plot. What do your plots show?
###b)	Use the Box-Cox procedure to find an appropriate power transformation by using $\lambda$ = -.2, -.1,0, .1, .2. What transformation of Y is suggested?
###c)	Use the transformation suggested by part b and obtain the estimated linear regression function for the transformed data.
###d)	Express the estimated regression function in the original units. Predict Y for X=0.5 and calculate the 95% confidence interval

### _Solution:_
### a)
```{r}
y<-c(243,195,275,190,213,249,239,243,269,273)
x<-c(5.5,5.3,5.6,5.3,5.4,5.5,5.5,5.5,5.6,5.6)
f5<-lm(y~x)
summary(f5)
ei.5<-f5$residuals
yhat.5=f5$fitted.values
stdei.5<- rstandard(f5)
par(mfrow=c(1,2))
plot(ei.5,yhat.5,xlab="Residuals",ylab="Fitted Values")
qqnorm(stdei.5,ylab="Standardized Residuals",xlab="Normal Scores", main="QQ Plot") 
qqline(stdei.5,col = "steelblue", lwd = 2)
```
QQ plot shows S pattern, indicates problem with linearity. Log transformation might be needed. box-cox transformation should be performed.

### b)

```{r}
library(MASS)
boxcox(f5,lamda=seq(-2,2,by=0.1))
```

boxcox plot shows that lamda is around zero, indicating log transformation
```{r}
y1<-log(y)
f6<-lm(y1~x)
summary(f6)
```


### d)
log(Y) = -0.9380 +  1.16903 X
Y= exp(-0.9380 +  1.16903 X)

```{r}
exp(predict.lm(f6,data.frame(x=0.5),interval = "confidence", level = 1-0.05))
```
